---
title: "04_eXtreme_Gradient_Boosting"
author: "EP"
date: "2026-07-19"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# eXtreme Gradient Boosting (XGBoost)

```{r message = FALSE}
library(targets)
library(tidymodels)
library(xgboost)
library(sf)
library(dplyr)

set.seed(399)
```

As seen in the Random Forest models, `n_days` is dominant predictor. For species richness, the full model has an excellent ROC AUC (0.956) dropping to 0.706 in the ND model (no `n_days`) which is weaker but moderately useable. However, Simpson diversity models are both weaker. The full model has a moderate ROC AUC of 0.712, dropping to a much weaker 0.662 in the ND model.
City level breakdown shows consistency between cities in the full richness model (ROC AUC ranging from 0.981-0.920), whereas the ND richness model has more variabilty (0.795-0.648). This suggests that without the dominant predictor, `n_days`, the model does not generalise to unseen cities as well. Both Simpson models also see greater variation in their ROC AUC scores: full (0.810-0.674) and ND (0.728-0.605). 
Comparison of the success rate vs prediction rate AUC in final full richness model shows a small decrease (success 0.985 vs prediction 0.956) suggesting the model fits the training data well and can predict unseen data well. However, when removing `n_days` there is a bigger gap (success 0.835 vs prediction 0.706) suggesting more overfitting now the model has to rely on weaker, potentially noisier, signals. Both the Simpson models have larger gaps: full (success 0.792 vs prediction 0.712) and ND (success 0.722 vs prediction 0.662).  
However, removing `n_days` highlights `weighted_temp`, `pct_greenspace`, and `weighted_precip` are drivers of richness diversity. `pct_greenspace` and `avg_area_ha` are top drivers of Simpson diversity. This is in agreement with the previous Random Forest results. Despite having lower AUC scores, the ND models are more meaningful when interpreting the ecological drivers of biodiversity.
 
## Load Data
```{r message=FALSE}
setwd("~/CSCT_Pipeline")

grid_model_all <- tar_read(grid_model_all)

df <- grid_model_all %>%
  sf::st_drop_geometry()

n <- nrow(df)
df$simpson_adj <- (df$simpson * (n - 1) + 0.5) / n

predictors <- c("n_days", "dtm_elev", "avg_imperv", "rofsw_risk", "rofrs_risk",
                "avg_area_ha", "treeht", "weighted_precip", "weighted_temp",
                "dist_to_river", "pct_greenspace")
```

## Model Setup
```{r}
#Classification

df <- df %>%
  mutate(
    richness_class = factor(if_else(sp_richness >= median(sp_richness, na.rm = TRUE),
                                     "high", "low"),
                             levels = c("low", "high")),
    simpson_class = factor(if_else(simpson_adj >= median(simpson_adj, na.rm = TRUE),
                                    "high", "low"),
                            levels = c("low", "high")))

#XGBoost Specification

xgb_spec_class <- boost_tree(
  trees          = tune(),
  tree_depth     = tune(),
  learn_rate     = tune(),
  loss_reduction = tune(),
  mtry           = tune(),
  min_n          = tune(),
  sample_size    = tune()) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

#Cross-validation folds
folds <- group_vfold_cv(df, group = city)

fold_city_lookup <- folds %>%
  mutate(city = purrr::map_chr(splits, ~ as.character(unique(assessment(.x)$city)))) %>%
  select(id, city)

#Recipes and workflows
richness_recipe_xgb <- recipe(
  as.formula(paste("richness_class ~", paste(predictors, collapse = " + "))),
  data = df) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_naomit(all_numeric_predictors(), all_outcomes()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())
 
richness_wf_xgb <- workflow() %>%
  add_recipe(richness_recipe_xgb) %>%
  add_model(xgb_spec_class)
 
simpson_recipe_xgb <- recipe(
  as.formula(paste("simpson_class ~", paste(predictors, collapse = " + "))),
  data = df) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_naomit(all_numeric_predictors(), all_outcomes()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())
 
simpson_wf_xgb <- workflow() %>%
  add_recipe(simpson_recipe_xgb) %>%
  add_model(xgb_spec_class)

```

## Tuning Setup
```{r}
xgb_grid <- grid_space_filling(
  trees(range = c(100, 1000)),
  tree_depth(range = c(2, 8)),
  learn_rate(range = c(-3, -0.5)),   # log10 scale: 0.001 - 0.316
  loss_reduction(range = c(-5, 1.5)),
  finalize(mtry(range = c(2, 8)), df[, predictors]),
  min_n(range = c(2, 15)),
  sample_prop(range = c(0.5, 1.0)),
  size = 30,
  type = "latin_hypercube")
```

## Hyperparameter Tuning
```{r}
richness_res_xgb <- tune_grid(
  richness_wf_xgb,
  resamples = folds,
  grid      = xgb_grid,
  metrics   = metric_set(roc_auc, accuracy, sensitivity, specificity))

simpson_res_xgb <- tune_grid(
  simpson_wf_xgb,
  resamples = folds,
  grid      = xgb_grid,
  metrics   = metric_set(roc_auc, accuracy, sensitivity, specificity))

autoplot(richness_res_xgb)
autoplot(simpson_res_xgb)
```

## Select Best
```{r}
show_best(richness_res_xgb, metric = "roc_auc", n = 5) %>%
  select(mtry, min_n, tree_depth, learn_rate, loss_reduction,
         sample_size, trees, mean, std_err) %>%
  mutate(across(c(mean, std_err), ~ round(.x, 3))) %>%
  rename("Mean ROC AUC" = mean, "Std. error" = std_err) %>%
  knitr::kable(caption = "Top 5 hyperparameter combinations: 
               species richness class (XGBoost, ranked by ROC AUC)")

show_best(simpson_res_xgb, metric = "roc_auc", n = 5) %>%
  select(mtry, min_n, tree_depth, learn_rate, loss_reduction,
         sample_size, trees, mean, std_err) %>%
  mutate(across(c(mean, std_err), ~ round(.x, 3))) %>%
  rename("Mean ROC AUC" = mean, "Std. error" = std_err) %>%
  knitr::kable(caption = "Top 5 hyperparameter combinations: 
               Simpson diversity class (XGBoost, ranked by ROC AUC)")

best_richness_xgb <- select_best(richness_res_xgb, metric = "roc_auc")
best_simpson_xgb  <- select_best(simpson_res_xgb,  metric = "roc_auc")
```

## Prediction Rate AUC
```{r}
overall_richness_xgb <- collect_metrics(richness_res_xgb) %>%
  filter(mtry == best_richness_xgb$mtry, min_n == best_richness_xgb$min_n,
         tree_depth == best_richness_xgb$tree_depth,
         learn_rate == best_richness_xgb$learn_rate,
         .metric == "roc_auc") %>%
  pull(mean)

overall_simpson_xgb <- collect_metrics(simpson_res_xgb) %>%
  filter(mtry == best_simpson_xgb$mtry, min_n == best_simpson_xgb$min_n,
         tree_depth == best_simpson_xgb$tree_depth,
         learn_rate == best_simpson_xgb$learn_rate,
         .metric == "roc_auc") %>%
  pull(mean)

city_richness_xgb <- collect_metrics(richness_res_xgb, summarize = FALSE) %>%
  filter(mtry == best_richness_xgb$mtry, min_n == best_richness_xgb$min_n,
         tree_depth == best_richness_xgb$tree_depth,
         learn_rate == best_richness_xgb$learn_rate,
         .metric == "roc_auc") %>%
  left_join(fold_city_lookup, by = "id") %>%
  select(city, .estimate) %>%
  rename(richness_auc = .estimate)

city_simpson_xgb <- collect_metrics(simpson_res_xgb, summarize = FALSE) %>%
  filter(mtry == best_simpson_xgb$mtry, min_n == best_simpson_xgb$min_n,
         tree_depth == best_simpson_xgb$tree_depth,
         learn_rate == best_simpson_xgb$learn_rate,
         .metric == "roc_auc") %>%
  left_join(fold_city_lookup, by = "id") %>%
  select(city, .estimate) %>%
  rename(simpson_auc = .estimate)

city_auc_xgb <- city_richness_xgb %>% full_join(city_simpson_xgb, by = "city")

overall_auc_xgb <- tibble(city = "Overall",
                           richness_auc = overall_richness_xgb,
                           simpson_auc  = overall_simpson_xgb)

classification_auc_table_xgb <- bind_rows(overall_auc_xgb, city_auc_xgb) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  rename("City (Held-Out)" = city,
         "Richness ROC AUC" = richness_auc,
         "Simpson ROC AUC"  = simpson_auc)

classification_auc_table_xgb %>%
  knitr::kable(caption = "Overall and city-level cross-validated ROC AUC for 
               species richness and Simpson diversity (XGBoost)")
```

## Final Model
```{r}
final_richness_wf_xgb <- finalize_workflow(richness_wf_xgb, best_richness_xgb)
richness_fit_xgb <- fit(final_richness_wf_xgb, data = df)

final_simpson_wf_xgb <- finalize_workflow(simpson_wf_xgb, best_simpson_xgb)
simpson_fit_xgb <- fit(final_simpson_wf_xgb, data = df)
```

## Success Rate AUC
```{r}
richness_success_xgb <- augment(richness_fit_xgb, new_data = df)
simpson_success_xgb  <- augment(simpson_fit_xgb,  new_data = df)

success_rate_richness_xgb <- roc_auc(richness_success_xgb, truth = richness_class,
                                      .pred_high, event_level = "second")
success_rate_simpson_xgb <- roc_auc(simpson_success_xgb, truth = simpson_class,
                                     .pred_high, event_level = "second")
```

## ROC Curves
```{r}
roc_curve(richness_success_xgb, truth = richness_class, .pred_high, event_level = "second")%>%
  autoplot() + labs(title = "Success rate curve: species richness (XGBoost)")

roc_curve(simpson_success_xgb, truth = simpson_class, .pred_high, event_level = "second")%>%
  autoplot() + labs(title = "Success rate curve: Simpson diversity (XGBoost)")
```

## Summary Table
```{r}
tibble::tibble(
  Outcome = c("Species richness", "Simpson diversity"),
  `Success rate AUC` = c(success_rate_richness_xgb$.estimate,
                          success_rate_simpson_xgb$.estimate),
  `Prediction rate AUC` = c(overall_richness_xgb, overall_simpson_xgb)
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable(caption = "XGBoost classification performance: 
               success rate vs. prediction rate AUC")
```

## Variable Importance 
```{r}
richness_fit_xgb %>%
  extract_fit_engine() %>%
  xgboost::xgb.importance(model = .) %>%
  as.data.frame() %>%
  ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col() +
  coord_flip() +
  labs(title = "Variable importance: species richness classification (XGBoost)",
       x = NULL, y = "Gain")
 
simpson_fit_xgb %>%
  extract_fit_engine() %>%
  xgboost::xgb.importance(model = .) %>%
  as.data.frame() %>%
  ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col() +
  coord_flip() +
  labs(title = "Variable importance: Simpson diversity classification (XGBoost)",
       x = NULL, y = "Gain")
```

# XGBoost Without `n_days`

## Set up and tuning 
```{r}

predictors_ND <- c("dtm_elev", "avg_imperv", "rofsw_risk", "rofrs_risk",
                "avg_area_ha", "treeht", "weighted_precip", "weighted_temp",
                "dist_to_river", "pct_greenspace")
#Recipes and workflows
richness_recipe_xgb_ND <- recipe(
  as.formula(paste("richness_class ~", paste(predictors_ND, collapse = " + "))),
  data = df) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_naomit(all_numeric_predictors(), all_outcomes()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())
 
richness_wf_xgb_ND <- workflow() %>%
  add_recipe(richness_recipe_xgb_ND) %>%
  add_model(xgb_spec_class)
 
simpson_recipe_xgb_ND <- recipe(
  as.formula(paste("simpson_class ~", paste(predictors_ND, collapse = " + "))),
  data = df) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_naomit(all_numeric_predictors(), all_outcomes()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())
 
simpson_wf_xgb_ND <- workflow() %>%
  add_recipe(simpson_recipe_xgb_ND) %>%
  add_model(xgb_spec_class)

#Hyperparameter Tuning
richness_res_xgb_ND <- tune_grid(
  richness_wf_xgb_ND,
  resamples = folds,
  grid      = xgb_grid,
  metrics   = metric_set(roc_auc, accuracy, sensitivity, specificity))

simpson_res_xgb_ND <- tune_grid(
  simpson_wf_xgb_ND,
  resamples = folds,
  grid      = xgb_grid,
  metrics   = metric_set(roc_auc, accuracy, sensitivity, specificity))

autoplot(richness_res_xgb_ND)
autoplot(simpson_res_xgb_ND)

#Select Best
show_best(richness_res_xgb_ND, metric = "roc_auc", n = 5) %>%
  select(mtry, min_n, tree_depth, learn_rate, loss_reduction,
         sample_size, trees, mean, std_err) %>%
  mutate(across(c(mean, std_err), ~ round(.x, 3))) %>%
  rename("Mean ROC AUC" = mean, "Std. error" = std_err) %>%
  knitr::kable(caption = "Top 5 hyperparameter combinations: 
               species richness ND class (XGBoost, ranked by ROC AUC)")

show_best(simpson_res_xgb_ND, metric = "roc_auc", n = 5) %>%
  select(mtry, min_n, tree_depth, learn_rate, loss_reduction,
         sample_size, trees, mean, std_err) %>%
  mutate(across(c(mean, std_err), ~ round(.x, 3))) %>%
  rename("Mean ROC AUC" = mean, "Std. error" = std_err) %>%
  knitr::kable(caption = "Top 5 hyperparameter combinations: 
               Simpson diversity ND class (XGBoost, ranked by ROC AUC)")

best_richness_xgb_ND <- select_best(richness_res_xgb_ND, metric = "roc_auc")
best_simpson_xgb_ND  <- select_best(simpson_res_xgb_ND,  metric = "roc_auc")
```

## Prediction Rate AUC
```{r}
overall_richness_xgb_ND <- collect_metrics(richness_res_xgb_ND) %>%
  filter(mtry == best_richness_xgb_ND$mtry, min_n == best_richness_xgb_ND$min_n,
         tree_depth == best_richness_xgb_ND$tree_depth,
         learn_rate == best_richness_xgb_ND$learn_rate,
         .metric == "roc_auc") %>%
  pull(mean)

overall_simpson_xgb_ND <- collect_metrics(simpson_res_xgb_ND) %>%
  filter(mtry == best_simpson_xgb_ND$mtry, min_n == best_simpson_xgb_ND$min_n,
         tree_depth == best_simpson_xgb_ND$tree_depth,
         learn_rate == best_simpson_xgb_ND$learn_rate,
         .metric == "roc_auc") %>%
  pull(mean)

city_richness_xgb_ND <- collect_metrics(richness_res_xgb_ND, summarize = FALSE) %>%
  filter(mtry == best_richness_xgb_ND$mtry, min_n == best_richness_xgb_ND$min_n,
         tree_depth == best_richness_xgb_ND$tree_depth,
         learn_rate == best_richness_xgb_ND$learn_rate,
         .metric == "roc_auc") %>%
  left_join(fold_city_lookup, by = "id") %>%
  select(city, .estimate) %>%
  rename(richness_auc = .estimate)

city_simpson_xgb_ND <- collect_metrics(simpson_res_xgb_ND, summarize = FALSE) %>%
  filter(mtry == best_simpson_xgb_ND$mtry, min_n == best_simpson_xgb_ND$min_n,
         tree_depth == best_simpson_xgb_ND$tree_depth,
         learn_rate == best_simpson_xgb_ND$learn_rate,
         .metric == "roc_auc") %>%
  left_join(fold_city_lookup, by = "id") %>%
  select(city, .estimate) %>%
  rename(simpson_auc = .estimate)

city_auc_xgb_ND <- city_richness_xgb_ND %>% full_join(city_simpson_xgb_ND, by = "city")

overall_auc_xgb_ND <- tibble(city = "Overall",
                           richness_auc = overall_richness_xgb_ND,
                           simpson_auc  = overall_simpson_xgb_ND)

classification_auc_table_xgb_ND <- bind_rows(overall_auc_xgb_ND, city_auc_xgb_ND) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  rename("City (Held-Out)" = city,
         "Richness ROC AUC" = richness_auc,
         "Simpson ROC AUC"  = simpson_auc)

classification_auc_table_xgb_ND %>%
  knitr::kable(caption = "Overall and city-level cross-validated ROC AUC for 
               ND species richness and Simpson diversity (XGBoost)")
```

## Final Model
```{r}
final_richness_wf_xgb_ND <- finalize_workflow(richness_wf_xgb_ND, best_richness_xgb_ND)
richness_fit_xgb_ND <- fit(final_richness_wf_xgb_ND, data = df)

final_simpson_wf_xgb_ND <- finalize_workflow(simpson_wf_xgb_ND, best_simpson_xgb_ND)
simpson_fit_xgb_ND <- fit(final_simpson_wf_xgb_ND, data = df)
```

## Success Rate AUC
```{r}
richness_success_xgb_ND <- augment(richness_fit_xgb_ND, new_data = df)
simpson_success_xgb_ND  <- augment(simpson_fit_xgb_ND,  new_data = df)

success_rate_richness_xgb_ND <- roc_auc(richness_success_xgb_ND, truth = richness_class,
                                      .pred_high, event_level = "second")
success_rate_simpson_xgb_ND <- roc_auc(simpson_success_xgb_ND, truth = simpson_class,
                                     .pred_high, event_level = "second")
```

## ROC Curves
```{r}
roc_curve(richness_success_xgb_ND, truth = richness_class, .pred_high, 
          event_level = "second") %>%
  autoplot() + labs(title = "Success rate curve: ND species richness (XGBoost)")

roc_curve(simpson_success_xgb_ND, truth = simpson_class, .pred_high, 
          event_level = "second") %>%
  autoplot() + labs(title = "Success rate curve: ND Simpson diversity (XGBoost)")
```

## Summary Table
```{r}
tibble::tibble(
  Outcome = c("ND Species richness", "ND Simpson diversity"),
  `Success rate AUC` = c(success_rate_richness_xgb_ND$.estimate,
                          success_rate_simpson_xgb_ND$.estimate),
  `Prediction rate AUC` = c(overall_richness_xgb_ND, overall_simpson_xgb_ND)
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable(caption = "XGBoost classification performance: 
               ND success rate vs. ND prediction rate AUC")
```

## Variable Importance 
```{r}
richness_fit_xgb_ND %>%
  extract_fit_engine() %>%
  xgboost::xgb.importance(model = .) %>%
  as.data.frame() %>%
  ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col() +
  coord_flip() +
  labs(title = "Variable importance: ND species richness classification (XGBoost)",
       x = NULL, y = "Gain")
 
simpson_fit_xgb_ND %>%
  extract_fit_engine() %>%
  xgboost::xgb.importance(model = .) %>%
  as.data.frame() %>%
  ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col() +
  coord_flip() +
  labs(title = "Variable importance: ND Simpson diversity classification (XGBoost)",
       x = NULL, y = "Gain")
```