---
title: "pipeline_report"
author: "EP"
date: "2026-08-04"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(targets)
library(dplyr)
library(tidyr)
library(sf)
library(ggplot2)
library(patchwork)
library(knitr)
```

```{r}
tar_load(grid_model_all)
tar_load(richness_models)
tar_load(simpson_models)
```

# Overview
```{r}
grid_model_all %>%
  sf::st_drop_geometry() %>%
  dplyr::group_by(city) %>%
  dplyr::summarise(
    n_cells         = dplyr::n(),
    mean_richness   = mean(sp_richness, na.rm = TRUE),
    mean_shannon    = mean(shannon, na.rm = TRUE),
    mean_simpson    = mean(simpson, na.rm = TRUE),
    total_records   = sum(n_records, na.rm = TRUE),
    total_survey_days = sum(n_days, na.rm = TRUE),
    .groups = "drop") %>%
  knitr::kable(digits = 2, caption = "Summary of diversity and recording effort by city")
```

# Grid Validation

## Grid geometry
```{r}
# Check all grid cells are valid and the correct size

geom_check <- grid_model_all %>%
  dplyr::mutate(
    valid_geom = sf::st_is_valid(.),
    area    = as.numeric(sf::st_area(.)) / 10000) %>%
  sf::st_drop_geometry() %>%
  dplyr::group_by(city) %>%
  dplyr::summarise(
    n_cells        = dplyr::n(),
    n_invalid_geom = sum(!valid_geom),
    mean_area   = mean(area),
    min_area    = min(area),
    max_area    = max(area),
    .groups = "drop")

geom_check %>%
  knitr::kable(digits = 2, caption = "Grid geometry check: validity and cell area by city")
```

```{r}
# Check the CRS is consistent
crs_list <- sf::st_crs(grid_model_all)$input
cat("CRS used across all cities:", crs_list, "\n")
```

## Check for duplicate and missing cells
```{r}
dup_check <- grid_model_all %>%
  sf::st_drop_geometry() %>%
  dplyr::count(city, cell_id, name = "n_occurrences") %>%
  dplyr::filter(n_occurrences > 1)

if (nrow(dup_check) == 0) {
  cat("No duplicate cell IDs found within any city.\n")
} else {
  dup_check %>% knitr::kable(caption = "Duplicate grid cells detected")
}
```

## Check for missing values
```{r}
predictor_cols <- c("sp_richness", "simpson", "shannon", "n_days", "n_records",
                     "dtm_elev", "avg_imperv", "avg_area_ha", "treeht",
                     "weighted_precip", "weighted_temp", "dist_to_river",
                     "pct_greenspace", "rofsw_risk", "rofrs_risk")

grid_model_all %>%
  sf::st_drop_geometry() %>%
  dplyr::select(city, dplyr::any_of(predictor_cols)) %>%
  dplyr::group_by(city) %>%
  dplyr::summarise(dplyr::across(everything(), ~ sum(is.na(.x))), .groups = "drop") %>%
  knitr::kable(caption = "Count of missing (NA) values per predictor, by city")
```

## Check that values are within a reasonable range
i.e. a percentage is between 0-100.
```{r}
range_check <- grid_model_all %>%
  sf::st_drop_geometry() %>%
  dplyr::summarise(
    dplyr::across(
      c(sp_richness, simpson, n_days, dtm_elev, avg_imperv,
        weighted_precip, weighted_temp, dist_to_river, pct_greenspace),
      list(min = ~ min(.x, na.rm = TRUE), max = ~ max(.x, na.rm = TRUE)),
      .names = "{.col}_{.fn}"))

range_check %>%
  tidyr::pivot_longer(everything(), names_to = "stat", values_to = "value") %>%
  knitr::kable(digits = 2, caption = "Min/max values for numerical variables")
```

## Cross-city consistency
```{r}
# Check that every city generated the same columns 

col_check <- grid_model_all %>%
  sf::st_drop_geometry() %>%
  split(.$city) %>%
  purrr::map(names)

all_same_cols <- length(unique(col_check)) == 1
cat("All cities have an identical column structure:", all_same_cols, "\n")
```

# Spatial patterns

## Species richness
```{r}
# Build one ggplot per city &and wrap them into a single image with patchwork.
make_city_maps <- function(data, fill_var, fill_label, ncol = 3) {
  cities <- sort(unique(data$city))
  plots <- lapply(cities, function(cty) {
    ggplot(dplyr::filter(data, city == cty)) +
      geom_sf(aes(fill = .data[[fill_var]]), colour = NA) +
      scale_fill_viridis_c(name = fill_label, na.value = "grey90") +
      labs(title = cty) +
      theme_minimal() +
      theme(
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        plot.title = element_text(size = 10, face = "bold"))})
  patchwork::wrap_plots(plots, ncol = ncol)}
```

```{r}
make_city_maps(grid_model_all, "sp_richness", "Richness")
```

## Simpson diversity

```{r}
make_city_maps(grid_model_all, "simpson", "Simpson")
```

# Flood Risk summary

```{r}
grid_model_all %>%
  sf::st_drop_geometry() %>%
  dplyr::count(city, rofsw_risk) %>%
  tidyr::pivot_wider(names_from = city, values_from = n, values_fill = 0) %>%
  knitr::kable(caption = "Cell counts by surface water flood risk (RoFSW) category and city")
```

```{r}
grid_model_all %>%
  sf::st_drop_geometry() %>%
  dplyr::count(city, rofrs_risk) %>%
  tidyr::pivot_wider(names_from = city, values_from = n, values_fill = 0) %>%
  knitr::kable(caption = "Cell counts by river and sea flood risk (RoFRS) category and city")
```

# Models

```{r}
# Manual likelihood ratio test: works for any models supporting logLik()/df.residual()
lr_test <- function(null_model, full_model) {
  ll_null <- as.numeric(logLik(null_model))
  ll_full <- as.numeric(logLik(full_model))
  df_null <- attr(logLik(null_model), "df")
  df_full <- attr(logLik(full_model), "df")

  stat <- 2 * (ll_full - ll_null)
  df   <- df_full - df_null
  p    <- pchisq(stat, df = df, lower.tail = FALSE)

  tibble::tibble(
    logLik_null = ll_null,
    logLik_full = ll_full,
    AIC_null    = AIC(null_model),
    AIC_full    = AIC(full_model),
    chisq       = stat,
    df          = df,
    p_value     = p)}
```

## Species richness (negative binomial GLM)

```{r}
summary(richness_models$full)$coefficients %>%
  as.data.frame() %>%
  tibble::rownames_to_column("term") %>%
  knitr::kable(digits = 3, caption = "Full richness model coefficients")
```

```{r}
lr_test(richness_models$null, richness_models$full) %>%
  knitr::kable(digits = 3, caption = "Richness: Null vs. Full models")
```

## Simpson diversity (beta regression)

```{r}
summary(simpson_models$full)$coefficients$mean %>%
  as.data.frame() %>%
  tibble::rownames_to_column("term") %>%
  knitr::kable(digits = 3, caption = "Full Simpson model coefficients")
```

```{r}
lr_test(simpson_models$null, simpson_models$full) %>%
  knitr::kable(digits = 3, caption = "Simpson: Null vs. Full Models")
```

# Recording effort

```{r}
grid_model_all %>%
  sf::st_drop_geometry() %>%
  ggplot(aes(x = n_days, y = sp_richness, colour = city)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Survey days per cell", y = "Species richness", title = "Sampling Effort by City") +
  theme_minimal()
```
